Partial rank data with the \pkg{hyper2} package: likelihood functions for generalized Bradley-Terry models

### 2019-05-07 inst/collusion_chess_draws.Rnw
work

% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
\documentclass[nojss]{jss}
\usepackage{dsfont}
\usepackage{bbm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{wasysym}
\usepackage{amssymb}
\usepackage{yfonts}
\usepackage{booktabs}

@@

Consider first the standard Bradley-Terry model for $n$ players.
First we ignore the effect of playing white.  We wish to find numbers
("strengths") $p_i,1\leqslant i\leqslant n$ with $p_i\geqslant 0, \sum
p_i=1$ such that the probability of $i$ beating $j\neq i$ is
$\frac{p_i}{p_i+p_j}$.  A suitable likelihood function, on the
assumption of independence of trials, would be

@@

which might be preferable on grounds of clarity.  We now apply this
likelihood model to a real dataset, shown in table~\ref{kka}.  The
likelihood function is easily maximized,
figure~\ref{maxp_karpov_kasparov_anand}.

\begin{figure}[htbp]
\begin{center}
<<rowing_maxp,echo=TRUE,fig=TRUE>>=
dotchart(maxp(karpov_kasparov_anand),pch=16,xlim=c(0,0.55))
@
\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
\label{maxp_karpov_kasparov_anand} under a likelihood defined as Equation~\ref{likelihood_kka}}
  \end{center}
\end{figure}

@@

Figure~\ref{kka} suggests a number of interesting hypotheses.
Firstly, we observe that $p_D=0$ cannot possibly be the case: the
likelihood would be identically zero, owing to the nonzero term in the
numerator of the last term of Equation~\ref{likelihood_kka}.  Indeed,
we can give a maximum likelihood estimate for the various strengths
using the \code{maxp()}``` functionality of the software:

@@

We see that $\widehat{p_D}\simeq 0.53$, compare the naive estimate of
$\frac{224}{365}\simeq 0.61$ given by observing 224 draws among 265
matches.  Secondly, we may make inferences about $p_W$.  It is
conceivable that $p_W=0$ as this term does not appear on its own in
the numerator of any term.  We may test $H_0\colon p_W=0$ using the
method of support.

<<>>=
small <- 1e-4

@@

that is, well over 7 units of support, exceeding Edwards's
two-units-of-support per degree of freedom and we can reject $p_W=0$.

One very natural hypothesis would be $H_{A=K=K}\colon
p_\mathrm{Karpov}=p_\mathrm{Kasparov}=p_\mathrm{Anand}$.  This too is
readily testable:

<<>>=
jj <- read.table("stockholm1962.txt",header=FALSE)
results <- as.matrix(jj[,-(1:2)])
players <- as.character(jj$V1)
nationality <- as.character(jj$V2)
rownames(results) <- players
colnames(results) <- substr(players,1,1)
results
@

(this taken from Wikipedia
\url{https://en.wikipedia.org/wiki/World_Chess_Championship_1963}).
In this dataset, "0" means a loss, "1" a draw and "2" a win.  Thus
Fisher (row 1) beat Korchnoi (column 4).  As a consistency check we
verify that the results table is antisymmetric:

<<>>=

all(t(results) == 2-results)
which(results != 2-t(results),arr.ind=TRUE)
@

Thus Barcza vs Portisch is problematic.  Glossing over this

>>>>

% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
\documentclass[nojss]{jss}
\usepackage{dsfont}
\usepackage{bbm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{wasysym}
\usepackage{amssymb}
\usepackage{yfonts}
\usepackage{booktabs}
\usepackage{units}
\usepackage{xcolor}

@@

Consider first the standard Bradley-Terry model for $n$ players.
First we ignore the effect of playing white.  We wish to find numbers
(``strengths") $p_i$ for $1\leqslant i\leqslant n$ with $p_i\geqslant 0, \sum p_i=1$ such that the probability of $i$ beating $j\neq i$ is $\frac{p_i}{p_i+p_j}$.  A suitable likelihood function, on the assumption of independence of trials, would be

@@

which might be preferable on grounds of clarity.  We now apply this
likelihood model to real datasets.

\section{Anand, Karpov, Kasparov}

Grandmasters Anand, Karpov, Kasparov have long-standing rivalry and
have played many matches amongst themselves.  A summary may be found
in Table~\ref{kka}.

@@

\begin{equation}\label{kka_likelihood}
\mathcal{L}\left(p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks},W,D\right)=
\frac{
p_\mathrm{An}^{15}
p_\mathrm{Kr}^{12} 
p_\mathrm{Ks}^{20}
D^{224}
\left(p_\mathrm{Kr} + W\right)^{25}
\left(p_\mathrm{Ks} + W\right)^{45}
\left(p_\mathrm{An} + W\right)^{24}
}{
\left(p_\mathrm{Kr} + p_\mathrm{Ks} + W + D\right)^{193}
\left(p_\mathrm{Kr} + p_\mathrm{An} + W + D\right)^{92}
\left(p_\mathrm{Ks} + p_\mathrm{An} + W + D\right)^{80}
}
\end{equation}

This likelihood function is easily maximized:

@@

\begin{figure}[htbp]
\begin{center}
<<ddc,echo=TRUE,fig=TRUE>>=
dotchart(maxp(karpov_kasparov_anand),pch=16,xlim=c(0,0.55))
@
\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
\label{maxp_karpov_kasparov_anand} under a likelihood defined as Equation~\ref{likelihood_kka}}
  \end{center}
\end{figure}

and this is show visually in figure~\ref{maxp_karpov_kasparov_anand},
which suggests a number of interesting and testable statistical
hypotheses.  Firstly, we observe that $D=0$ cannot possibly be the
case: the likelihood would be identically zero, owing to the nonzero
term in the numerator of the last term of
Equation~\ref{likelihood_kka}.  However, we see that the maximum
likelihood estimate for Draw strength $\hat{D}\simeq 0.53$: compare
the na\"ive estimate of $\frac{224}{365}\simeq 0.61$ given by
observing 224 draws among 265 matches.  The MLE is lower because the
maximization process automatically adjusts for the different players'
strengths.  Secondly, we may make inferences about $W$, the extra
strength conferred by playing White: it is conceivable that $W=0$ as
this term does not appear on its own in the numerator of any term.  We
may test $H_0\colon W=0$ using the method of support by maximizing the
likelihood subject the constraint $W=0$ and comparing with the
likelihood when the optimization is free.

<<echo=FALSE>>=
small <- 1e-4

@@

The result of numerical optimization is 7.304, that is, well over the
two-units-of-support per degree of freedom suggested by Edwards; we
can convincingly reject the hypothesis that $W=0$ and infer that
playing white confers a significant advantage.  Of course, this is
nothing new CITATION NEEDED, but it is now possible to go
significantly beyond mere existence of the effect and to quantify it.
Taking the MLE above, and considering a match between say Karpov and
Anand, the strengths are Karpov=0.064, Anand=0.151, white =0.104 and
draw =0.53.  Thus we would estimate that Karpov playing white against
Kasparov would win with probability
$\frac{0.064+0.104}{0.064+0.104+0.148+0.532}\simeq 0.198$, draw with
probability $\frac{532}{64+104+148+532}\simeq 0.625$, and lose with
probability $\simeq 0.174$ (compare $+0.075\,=0.627-0.297$ playing
black).

One very natural hypothesis would be that the players are of equal
strength and we migth write $H_=\colon
p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.  This too is readily
testable by the device of calculating the maximum support over the
region in which $p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.

<<print=FALSE,echo=FALSE>>=
objective <- function(V){
  x <- V[1]  # strength of all three players
  w <- V[2]  # white player advantage
  D <- 1-3*x-w  # Draw monster
  return(  
      -loglik(karpov_kasparov_anand,c(x,x,x,w))  # negative as we minimize
  )
}

jj <-
constrOptim(
    theta=c(0.1,0.4),
    f=objective,
    grad=NULL,
    ui = rbind(diag(2),-diag(2),c(-3,-1)),
    ci = c(0,0,-1,-1,-1)
)

unconstrained <- maxp(karpov_kasparov_anand,give=TRUE)$value
constrained <- jj$value

extra_support <-   unconstrained + constrained  # negative used

pchisq(2*extra_support,df=2,lower.tail=FALSE)
@

In this case the extra support in going from the constrained to the
free optimization is about 4.93.  However, this time the constraint
corresponds to a loss of {\em two} degrees of freedom so we would need
4 units of support, which is attained in this case.  Alternatively we
might follow Wilks, and observe that the value of
$2\Lambda=2*4.92=9.84$ lies in the tail region of its asymptotic
distribution with a p-value of about 0.0072.  We can justifiably
assert that the players do indeed have different strengths, even if we
account for white advantage and the proclivity to draw.

\subsection{Each player has a draw monster}

Players with a cautious style tend to draw CITATION NEEDED and this is
an important and interesting component of the game.  To investigate
this we need a likelihood function that can account for possible
differences in draw proclivity among the players.  This may be
achieved by considering a match between two players to be a contest
between five entities: the two real players, the white advantage, and
two personalised draw entities, one per real player; a drawn match is
counted as a win for either of the draw ghosts.  Under these
assumptions, table~\ref{kka} translates into the following likelihood
function:

\begin{figure}[htbp]
\begin{center}
<<ddcc,echo=TRUE,fig=TRUE>>=
pie(maxp(kka_3draws))
@
\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
under a likelihood function that admits differential proclivities to draw}
  \end{center}
\end{figure}

\begin{equation}
\mathcal{L}\left(
p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks}
,W,
d_\mathrm{An},d_\mathrm{Kr},d_\mathrm{Ks}
\right)=
\frac{
\parbox{3in}{$
p_\mathrm{An}^{15}
p_\mathrm{Kr}^{12}
p_\mathrm{Ks}^{20}
\left(p_\mathrm{Kr} + W\right)^{25}
\left(p_\mathrm{Ks} + W\right)^{45} 
\left(p_\mathrm{An} + W\right)^{24}\\ \rule{10mm}{0mm}
\left(d_\mathrm{Kr} + d_\mathrm{Ks}\right)^{129}
\left(d_\mathrm{Kr} + d_\mathrm{An}\right)^{49}
\left(d_\mathrm{Ks} + d_\mathrm{An}\right)^{46}
$}}{\parbox{3in}{$
\left(p_\mathrm{Kr} + p_\mathrm{Ks} + W + d_\mathrm{Kr} + d_\mathrm{Ks}\right)^{193}\\ \rule{10mm}{0mm}
\left(p_\mathrm{Kr} + p_\mathrm{An} + W + d_\mathrm{Kr} + d_\mathrm{An}\right)^{92}\\ \rule{20mm}{0mm}
\left(p_\mathrm{Ks} + p_\mathrm{An} + W + d_\mathrm{Ks} + d_\mathrm{An}\right)^{80}
$}}
\end{equation}

<<>>=
round(maxp(kka_3draws),3)
@

We see considerable variability in proclivity to draw and again this
may be tested with the Method of Support, which gives a $p$-value of
about 0.039, borderline significant: an interesting result, given that
in this case no
player's draw monster appears alone anywhere in the likelihood function.

<<echo=FALSE,print=FALSE>>=
objective <- function(V){
  karpov <- V[1]
  kasparov <- V[2]
  anand <- V[3]
  white <- V[4]

  draw <- (1-karpov-kasparov-anand-white)/3

    return(  
        -loglik(kka_3draws,
                c(karpov,kasparov,anand,white,draw,draw)
                )
        )
}

jj <-
constrOptim(
    theta=c(0.1,0.1,0.1,0.1),
    f=objective,
    grad=NULL,
    ui <- rbind(
        diag(nrow=4),
        c(-1,-1,-1,-1)
        ),
    ci = c(
        rep(0,4),
        -1
        )
)

unconstrained <- maxp(kka_3draws,give=TRUE)$value
constrained <- jj$value
extra_support <-   unconstrained + constrained
pchisq(2*extra_support,df=2,lower.tail=FALSE)
@

\section{Chess collusion}

\newcommand{\halfw}{$\nicefrac{1}{2}$}
\newcommand{\halfb}{\colorbox{black}{\textcolor{white}{$\nicefrac{{\bf 1}}{{\bf 2}}$}}}
\newcommand{\zerow}{$0$}
\newcommand{\zerob}{\colorbox{black}{\textcolor{white}{$0$}}}
\newcommand{\oneww}{$1$}
\newcommand{\onebb}{\colorbox{black}{\textcolor{white}{$1$}}}

\begin{table}
\centering
\begin{tabular}{l|cccccc}
& \rotatebox{-90}{Fischer} 
&\rotatebox{-90}{Geller} 
&\rotatebox{-90}{Petrosian\ }
&\rotatebox{-90}{Korchnoi} 
&\rotatebox{-90}{Filip}
& $\cdots$
\\ \hline %       Fischer   Geller    Petrosian Korchnoi    Filip
Fischer     (USA)  &   -     & \halfb  & \halfw  & \oneww    & \halfb   &\\
Geller    (USSR)  & \halfw  &    -    & \halfb  & \halfw    & \halfw   &\\
Petrosian (USSR)  & \halfb  & \halfw  &    -    & \halfb    & \halfw   &\\
Korchnoi  (USSR)  & \zerob  & \halfb  & \halfw  &   -       & \oneww   &\\
Filip      (TCH)  & \halfw  & \halfb  & \halfb  & \zerob    &  -       &\\
$\vdots$&&&&&\\
\end{tabular}\label{interzonal}
\caption{Subset of the Fifth Interzonal round-robin, Stockholm 1962.
  Results (0,\halfw,1) and box colour refer to row player.  Thus, for
  example, Fischer played black against Geller and drew}
\end{table}

For the results shown in table~\ref{interzonal}, we would have

\begin{equation}
\mathcal{L}\left(F,G,P,K,L,W,D\right)=
%                D/(F+G+W+D)  (F+P+D)/(F+P+D+W)   (F+W)/(F+W+D+K)   D/(F+W+L+D)
%                             D/(G+P+D+W)         D/(G+K+W+D)       D/(G+F+W+D)
%                                                 D/(P+K+W+D)       D/(P+L+W+D)
%                                                                   (K+W)/(K+L+W+D)
\frac{
  D^8(F+W)(K+W)
}{\parbox{4in}{$
    (F+G+W+D)(F+P+W+D)\\ \rule{10mm}{0mm}(F+K+W+D)  (F+L+W+D)\\ \rule{20mm}{0mm}
         (G+P+W+D)  (G+K+W+D)\\ \rule{30mm}{0mm}(G+F+W+D)
                    (P+K+W+D)\\ \rule{40mm}{0mm}(P+L+W+D)
                               (K+L+W+D)
 $}}
\end{equation}

\noindent (the numerator includes 8 draws and 2 wins, one for Fischer
and one for Korchnoi, both playing White); but the full likelihood
function for the interzonal has 297 terms, including $D^{110}$.  To
investigate the possibility of Soviet collusion we introduce a new
draw entity: a soviet Draw monster $D_s$, who wins when two Soviet
players draw, and is inactive whenever a game with at least one
non-Soviet player is drawn.  Then $H_N\colon D_s=D$ corresponds to no
collusion and the likelihood function for table~\ref{interzonal} would
be

\begin{equation}\label{collusion}
\mathcal{L}\left(F,G,P,K,L,W,D,D_s\right)=
%                D/(F+G+W+D)  (F+P+D)/(F+P+D+W)   (F+W)/(F+W+D+K)   D/(F+W+L+D)
%                             D/(G+P+D+W)         D/(G+K+W+D)       D/(G+F+W+D)
%                                                 D/(P+K+W+D)       D/(P+L+W+D)
%                                                                   (K+W)/(K+L+W+D)
\frac{
  D_s^3D^5(F+W)(K+W)
}{\parbox{4in}{$
    (F+G+W+D)(F+P+W+D)\\ \rule{10mm}{0mm}(F+K+W+D)  (F+L+W+D)\\ \rule{20mm}{0mm}
         (G+P+W+D_s)  (G+K+W+D_s)\\ \rule{30mm}{0mm}(G+F+W+D)
                    (P+K+W+D_s)\\ \rule{40mm}{0mm}(P+L+W+D)
                               (K+L+W+D)
 $}}
\end{equation}

\begin{figure}[htbp]
\begin{center}
<<collusion,echo=TRUE,fig=TRUE>>=
dotchart(maxp(collusion),pch=16)
@
\caption{sds\label{collusion_fig}}
  \end{center}
\end{figure}

In figure~\ref{collusion_fig}, note the high estimate for $D_s$ (also
note
Fischer's dominance).  The maximum likelihood estimate is $D\simeq 0.043, D_s=0.354$; a constrained maximization gives a likelihood ratio of about 5096, or a $p$-value of about $3.6\times 10^{-5}$.  Indeed we could go further and plot a profile likelihood curve for the log-odds ratio $\ln\left(D_s/D\right)$ which, if Dirichlet, would be approximately Gaussian CITEKENDALANDSTUARTHERE.  This is illustrated in Figure~\ref{DOESNOTEXIST}, that shows a credible interval of about HAVENOTDONETHISET.
